---
title: "Restitution"
output: html_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###IMPORTS
```{r}
library(KMsurv)
library(survival)
library(ggplot2)
library(survminer)
library(data.table)
library(joineRML)
library(tidyr)
library(MASS)
```


```{r}
library(tidyr)
```

#setup
```{r}
set.seed(0)
data(pbc2)
```

###PART A

##Missing Values
```{r}
colSums(is.na(pbc2))
missing_row <- data.frame(id= pbc2$id, missing = rowSums(is.na(pbc2)))
missing_row <-missing_row[missing_row$missing>0,]
```

##Constant variables
```{r}
dif_values <- data.table(pbc2)[, lapply(.SD, function(x) length(unique(x))), by=id]
#^for each individual we check how many different values for each variable
check_constants <- colSums(dif_values[,!'id'])
check_constants #if the value is 312, the variable is constant for a given individual
#constant variables are: years, status, drug, age, sex, status2
pbc2_constants <- unique(pbc2[,c('years', 'status', 'drug', 'age', 'sex', 'status2')])
hist(pbc2$years)
barplot(prop.table(table(pbc2_constants$status)))
barplot(prop.table(table(pbc2_constants$drug)))
hist(pbc2$age)
barplot(prop.table(table(pbc2_constants$sex)))
barplot(prop.table(table(pbc2_constants$status2)))
```
## Quantitative variables
```{r}
plot(pbc2$year, pbc2$serBilir)
abline(lm(pbc2$serBilir ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$serChol)
abline(lm(pbc2$serChol ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$alkaline)
abline(lm(pbc2$alkaline ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$SGOT)
abline(lm(pbc2$SGOT ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$platelets)
abline(lm(pbc2$platelets ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$prothrombin)
abline(lm(pbc2$prothrombin ~ pbc2$year), col = "red")
```

```{r}
ggplot(pbc2[c('id','serBilir','year')], aes(year, serBilir, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','serChol','year')], aes(year, serChol, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','alkaline','year')], aes(year, alkaline, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','SGOT','year')], aes(year, SGOT, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','platelets','year')], aes(year, platelets, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','prothrombin','year')], aes(year, prothrombin, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
```

###PART B
##a)

```{r}
surv_spiders <- survfit(Surv(year,status2)~spiders, data=pbc2)
summary(surv_spiders)
ggsurvplot(surv_spiders)
```
##b)
```{r}
print(surv_spiders)
quantile(surv_spiders, prob=0.5)
quantile(surv_spiders, prob=0.25)
quantile(surv_spiders, prob=0.75)
```
##c)
```{r}
survdiff(Surv(year,status2)~spiders, data=pbc2) #log ranks
survdiff(Surv(year,status2)~spiders, data=pbc2, rho=1) #peto-peto weights
```

##d)
```{r}
cox_spiders <- coxph(Surv(year,status2)~spiders, data=pbc2)
summary(cox_spiders)
```
###PART C
##a)
```{r}
cox_all <- coxph(Surv(year,status2)~drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + serChol + albumin + alkaline + SGOT + platelets + prothrombin, data = pbc2)
cox_all
pbc2_no_na <- na.omit(pbc2)
cox_select <- stepAIC(coxph(Surv(year,status2)~drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + serChol + albumin + alkaline + SGOT + platelets + prothrombin, data = pbc2_no_na),
        scope = list(upper = ~., lower= ~spiders), 
        direction = 'both')
cox_select
```

##b)
```{r}
#
sum_cox_select <- summary(cox_select)
sum_cox_select$conf.int
```

(1) Dickson ER, Grambsch PM, Fleming TR, Fisher LD, Langworthy A. Prognosis in primary biliary cirrhosis: model for decision making. Hepatology. 1989 Jul;10(1):1-7. doi: 10.1002/hep.1840100102. PMID: 2737595.

(2) Murtaugh PA, Dickson ER, Van Dam GM, Malinchoc M, Grambsch PM, Langworthy AL, Gips CH. Primary biliary cirrhosis: prediction of short-term survival based on repeated patient visits. Hepatology. 1994 Jul;20(1 Pt 1):126-34. doi: 10.1016/0270-9139(94)90144-9. PMID: 8020881.