---
title: "Restitution"
output: html_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###IMPORTS
```{r}
library(KMsurv)
library(survival)
library(graphics)
library(ggplot2)
library(survminer)
library(survival)
library(data.table)
library(joineRML)
library(tidyr)
library(dplyr)
library(purrr)
library(mice)
library(zoo)
library(MASS) 
```


```{r}
library(tidyr)
```

#setup
```{r}
set.seed(0)
data(pbc2)
```

###PART A

##Missing Values
```{r}
colSums(is.na(pbc2))
missing_row <- data.frame(id= pbc2$id, missing = rowSums(is.na(pbc2)))
missing_row <-missing_row[missing_row$missing>0,]
```

##Constant variables
```{r}
dif_values <- data.table(pbc2)[, lapply(.SD, function(x) length(unique(x))), by=id]
#^for each individual we check how many different values for each variable
check_constants <- colSums(dif_values[,!'id'])
check_constants #if the value is 312, the variable is constant for a given individual
#constant variables are: years, status, drug, age, sex, status2
pbc2_constants <- unique(pbc2[,c('years', 'status', 'drug', 'age', 'sex', 'status2')])
hist(pbc2$years)
barplot(prop.table(table(pbc2_constants$status)))
barplot(prop.table(table(pbc2_constants$drug)))
hist(pbc2$age)
barplot(prop.table(table(pbc2_constants$sex)))
barplot(prop.table(table(pbc2_constants$status2)))
```
## Quantitative variables
```{r}
plot(pbc2$year, pbc2$serBilir)
abline(lm(pbc2$serBilir ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$serChol)
abline(lm(pbc2$serChol ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$alkaline)
abline(lm(pbc2$alkaline ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$SGOT)
abline(lm(pbc2$SGOT ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$platelets)
abline(lm(pbc2$platelets ~ pbc2$year), col = "red")
plot(pbc2$year, pbc2$prothrombin)
abline(lm(pbc2$prothrombin ~ pbc2$year), col = "red")
```

```{r}
ggplot(pbc2[c('id','serBilir','year')], aes(year, serBilir, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','serChol','year')], aes(year, serChol, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','alkaline','year')], aes(year, alkaline, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','SGOT','year')], aes(year, SGOT, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','platelets','year')], aes(year, platelets, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
ggplot(pbc2[c('id','prothrombin','year')], aes(year, prothrombin, group = id, color = id)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none")
```
```{r}
#HANDLING MISSING VALUES
# Step 1: Remove serChol
pbc2_locf <- pbc2 %>%
  select(-serChol)

# Step 2: Identify variables to impute (exclude id and year)
vars_to_impute <- setdiff(names(pbc2_locf), c("id", "year"))

# Step 3: Apply LOCF imputation grouped by patient ID
pbc2_locf <- pbc2_locf %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(across(all_of(vars_to_impute), ~ na.locf(.x, na.rm = FALSE))) %>%
  ungroup()

colSums(is.na(pbc2_locf))

sum(is.na(pbc2_locf$platelets)) 
pbc2_locf %>% filter(is.na(platelets)) %>% select(id, year)

# For these 4 observation a median imputation grou[ed by sex and drug group 
baseline <- pbc2_locf %>% filter(year == 0)

# Step 2: Compute median platelets per drug × sex group (excluding NAs)
median_by_group <- baseline %>%
  filter(!is.na(platelets)) %>%
  group_by(drug, sex) %>%
  summarise(median_platelets = median(platelets), .groups = "drop")

# Step 3: Join and impute missing baseline values
baseline_imputed <- baseline %>%
  left_join(median_by_group, by = c("drug", "sex")) %>%
  mutate(platelets = if_else(is.na(platelets), median_platelets, platelets)) %>%
  select(-median_platelets)

# Step 4: Replace the original baseline data in the full dataset
pbc2_filled <- pbc2_locf %>%
  filter(year > 0) %>%
  bind_rows(baseline_imputed) %>%
  arrange(id, year)

sum(is.na(pbc2_filled$platelets))  # should be 0
```


###PART B
##a)

```{r}
surv_spiders <- survfit(Surv(years,status2)~spiders, data=pbc2_filled)
summary(surv_spiders)
ggsurvplot(surv_spiders)
```
##b)
```{r}
print(surv_spiders)
quantile(surv_spiders, prob=0.5)
quantile(surv_spiders, prob=0.25)
quantile(surv_spiders, prob=0.75)
```
##c)
```{r}
survdiff(Surv(years,status2)~spiders, data=pbc2_filled) #log ranks
survdiff(Surv(years,status2)~spiders, data=pbc2_filled, rho=1) #peto-peto weights
```

##d)
```{r}
cox_spiders <- coxph(Surv(years,status2)~spiders, data=pbc2_filled)
summary(cox_spiders)
```
###PART C
##a)

```{r}
cox_all <- coxph(Surv(years,status2)~drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT + platelets + prothrombin, data = pbc2_filled)
cox_all
cox_select <- stepAIC(coxph(Surv(years,status2)~drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT + platelets + prothrombin, data = pbc2_filled),
        scope = list(upper = ~., lower= ~spiders), 
        direction = 'both')
cox_select
```

```{r}
#confidence intervals for the selected model
sum_cox_select <- summary(cox_select)
sum_cox_select$conf.int
```
```{r}
#local test of significance of spiders
linearHypothesis(cox_select, c("spidersYes"))
```
```{r}
#see practice session n°2, exercise 2, question 5 (timecode of recording 1h25min)
#first those fit are the estimates of the survival curves on subsets depending on our binary variable spiders
fitYes <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
              data = subset(pbc2_filled, spiders =='Yes')))
fitNo <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
              data = subset(pbc2_filled, spiders =='No')))

#surfit fuction provides values only at times of observations, which are different for the two subsets, so we build a time scale with all time points
cumhaz <- data.frame(time = sort(c(fitYes$time, fitNo$time)))

cumhazYes <- data.frame(time = fitYes$time, cumhaz = fitYes$cumhaz)
cumhazNo <- data.frame(time = fitNo$time, cumhaz = fitNo$cumhaz)
#on unobserved times, the last observation is carried forward
cumhaz$chSpidersYes <- cumhazYes$cumhaz[na.locf(match(cumhaz$time,cumhazYes$time), na.rm = FALSE)]
cumhaz$chSpidersNo <- cumhazNo$cumhaz[na.locf(match(cumhaz$time,cumhazNo$time), na.rm = FALSE)]
```


```{r}
attach(cumhaz)

plot(time, chSpidersYes, type = "s", ylab = "Cumulative hazards", xlab = "Time", col='red')
lines(time, chSpidersNo, type = "s", lty= 2, col = 'darkgreen')
legend(x=10, y = 0.1, legend = c("Spiders = Yes", "Spiders = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chSpidersYes), type = "s", ylab = "log(cumulative hazards)", xlab = "Time", col = 'red')
lines(time, log(chSpidersNo), type = "s", lty= 2, 'col' = 'darkgreen')
legend(x=10, y = -5, legend = c("Spiders = Yes", "Spiders = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chSpidersYes) - log(chSpidersNo), type = "s", ylab = "log(ratio cumulative hazard)", xlab = "Time")

plot(chSpidersNo, chSpidersYes, type = "s", ylab = "Cumulative hazard for spiders=Yes", xlab = "Cumulative hazard for spiders=No")
```

```{r}
fitMen <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
              data = subset(pbc2_filled, sex =='male')))
fitWomen <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
              data = subset(pbc2_filled, sex =='female')))

#surfit fuction provides values only at times of observations, which are different for the two subsets, so we build a time scale with all time points
cumhazSex <- data.frame(time = sort(c(fitMen$time, fitWomen$time)))

cumhazMen <- data.frame(time = fitMen$time, cumhazSex = fitMen$cumhaz)
cumhazWomen <- data.frame(time = fitWomen$time, cumhazSex = fitWomen$cumhaz)
#on unobserved times, the last observation is carried forward
cumhazSex$chMen <- cumhazMen$cumhazSex[na.locf(match(cumhazSex$time,cumhazMen$time), na.rm = FALSE)]
cumhazSex$chWomen <- cumhazWomen$cumhazSex[na.locf(match(cumhazSex$time,cumhazWomen$time), na.rm = FALSE)]
attach(cumhazSex)

plot(time, chMen, type = "s", ylab = "Cumulative hazards", xlab = "Time", col='red')
lines(time, chWomen, type = "s", lty= 2, col = 'darkgreen')
legend(x=0, y = 1.5, legend = c("Men", "Women"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chMen), type = "s", ylab = "log(cumulative hazards)", xlab = "Time", col = 'red')
lines(time, log(chWomen), type = "s", lty= 2, 'col' = 'darkgreen')
legend(x=10, y = -4, legend = c("Men", "Women"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chMen) - log(chWomen), type = "s", ylab = "log(ratio cumulative hazard)", xlab = "Time")

plot(chMen, chWomen, type = "s", ylab = "Cumulative hazard for men", xlab = "Cumulative hazard for women")
```

```{r}
#diagnostic plots for ascites
fitYes <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + spiders + hepatomegaly + ascites + edema + serBilir + albumin + alkaline + SGOT,
                        data = subset(pbc2_filled, ascites =='Yes')))
fitNo <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + spiders + hepatomegaly + ascites + edema + serBilir + albumin + alkaline + SGOT,
                       data = subset(pbc2_filled, ascites =='No')))

#surfit fuction provides values only at times of observations, which are different for the two subsets, so we build a time scale with all time points
cumhaz <- data.frame(time = sort(c(fitYes$time, fitNo$time)))

cumhazYes <- data.frame(time = fitYes$time, cumhaz = fitYes$cumhaz)
cumhazNo <- data.frame(time = fitNo$time, cumhaz = fitNo$cumhaz)
#on unobserved times, the last observation is carried forward
cumhaz$chAscitesYes <- cumhazYes$cumhaz[na.locf(match(cumhaz$time,cumhazYes$time), na.rm = FALSE)]
cumhaz$chAscitesNo <- cumhazNo$cumhaz[na.locf(match(cumhaz$time,cumhazNo$time), na.rm = FALSE)]

attach(cumhaz)

plot(time, chAscitesYes, type = "s", ylab = "Cumulative hazards", xlab = "Time", col='red')
lines(time, chAscitesNo, type = "s", lty= 2, col = 'darkgreen')
legend(x=1, y = 1.2, legend = c("Ascites = Yes", "Ascites = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chAscitesYes), type = "s", ylab = "log(cumulative hazards)", xlab = "Time", col = 'red')
lines(time, log(chAscitesNo), type = "s", lty= 2, 'col' = 'darkgreen')
legend(x=10, y = -4, legend = c("Ascites = Yes", "Ascites = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chAscitesYes) - log(chAscitesNo), type = "s", ylab = "log(ratio cumulative hazard)", xlab = "Time")

plot(chAscitesNo, chAscitesYes, type = "s", ylab = "Cumulative hazard for ascites=Yes", xlab = "Cumulative hazard for ascites=No")
```

```{r}
fitYes <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + spiders + hepatomegaly + ascites + edema + serBilir + albumin + alkaline + SGOT,
                        data = subset(pbc2_filled, hepatomegaly =='Yes')))
fitNo <- survfit(coxph(Surv(years, status2) ~ drug + age + sex + spiders + hepatomegaly + ascites + edema + serBilir + albumin + alkaline + SGOT,
                       data = subset(pbc2_filled, hepatomegaly =='No')))

#surfit fuction provides values only at times of observations, which are different for the two subsets, so we build a time scale with all time points
cumhaz <- data.frame(time = sort(c(fitYes$time, fitNo$time)))

cumhazYes <- data.frame(time = fitYes$time, cumhaz = fitYes$cumhaz)
cumhazNo <- data.frame(time = fitNo$time, cumhaz = fitNo$cumhaz)
#on unobserved times, the last observation is carried forward
cumhaz$chHepatomegalyYes <- cumhazYes$cumhaz[na.locf(match(cumhaz$time,cumhazYes$time), na.rm = FALSE)]
cumhaz$chHepatomegalyNo <- cumhazNo$cumhaz[na.locf(match(cumhaz$time,cumhazNo$time), na.rm = FALSE)]

attach(cumhaz)

plot(time, chHepatomegalyYes, type = "s", ylab = "Cumulative hazards", xlab = "Time", col='red')
lines(time, chHepatomegalyNo, type = "s", lty= 2, col = 'darkgreen')
legend(x=1, y = 1.1, legend = c("Hepatomegaly = Yes", "Hepatomegaly = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chHepatomegalyYes), type = "s", ylab = "log(cumulative hazards)", xlab = "Time", col = 'red')
lines(time, log(chHepatomegalyNo), type = "s", lty= 2, 'col' = 'darkgreen')
legend(x=7, y = -5, legend = c("Hepatomegaly = Yes", "Hepatomegaly = No"), col = c("darkgreen", 'red'), lty = c(2,1))

plot(time, log(chHepatomegalyYes) - log(chHepatomegalyNo), type = "s", ylab = "log(ratio cumulative hazard)", xlab = "Time")

plot(chHepatomegalyNo, chHepatomegalyYes, type = "s", ylab = "Cumulative hazard for hepatomegaly=Yes", xlab = "Cumulative hazard for hepatomegaly=No")
```


##b) Parametric regression models
```{r}
#see exercise session 2 recording at 2:19:30
fit_weibull <- survreg(Surv(years, status2)~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
        data = pbc2,
        dist = 'weibull')
fit_exp <- survreg(Surv(years, status2)~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
        data = pbc2,
        dist = 'exponential')
fit_loglog <- survreg(Surv(years, status2)~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
        data = pbc2,
        dist = 'loglogistic')
fit_lognorm <- survreg(Surv(years, status2)~ drug + age + sex + ascites + hepatomegaly + spiders + edema + serBilir + albumin + alkaline + SGOT,
        data = pbc2,
        dist = 'lognormal')
```

```{r}
#Model selection on AIC criteria
AIC <- c(extractAIC(fit_weibull)[2],extractAIC(fit_exp)[2],extractAIC(fit_loglog)[2],extractAIC(fit_lognorm)[2])
names(AIC) <- c('weibul','exponential', 'loglogistic', 'lognormal')
AIC
#lowest AIC is loglogistic, so it should be the one selected
```
```{r}
#specifications of selected model
summary(fit_loglog)
```





(1) Dickson ER, Grambsch PM, Fleming TR, Fisher LD, Langworthy A. Prognosis in primary biliary cirrhosis: model for decision making. Hepatology. 1989 Jul;10(1):1-7. doi: 10.1002/hep.1840100102. PMID: 2737595.

(2) Murtaugh PA, Dickson ER, Van Dam GM, Malinchoc M, Grambsch PM, Langworthy AL, Gips CH. Primary biliary cirrhosis: prediction of short-term survival based on repeated patient visits. Hepatology. 1994 Jul;20(1 Pt 1):126-34. doi: 10.1016/0270-9139(94)90144-9. PMID: 8020881.